Report on the Third Workshop on Sustainable Software for Science: Practice and Experiences (WSSSPE3)

### 2015-10-24 WSSSPE3/report/WSSSPE3_review.tex
Section 3: Keynote first draft completed.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Keynote} \label{sec:keynote} 
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\todo{need to update for WSSSPE3. 
Lead: S.-C. Choi. 
PDF slides at \url{http://tinyurl.com/qbbqgsj}.
Youtube Video at \url{http://tinyurl.com/q45kfcn}.
}

@@

Turk reminded the audience that the volunteer model means unpaid labor. On this
note, he recommended Ashe Dryden's blog
post\footnote{http://www.ashedryden.com/blog/the-ethics-of-unpaid-labor-and-the-oss-community}
on the ethics of unpaid labor and the open source software community. Often
times, a person funded to work full time on some scientific project can spend a
small amount of time for working on a piece of software necessary for the
project. However, all researchers' ability to participate in that volunteer
community are not always the same and may not always be aligned with the
research projects. From Turk's experience, we cannot always rely on unpaid labor
and volunteer time to sustain a piece of software---this came down to the
notions of the top-down and the bottom-up approaches, the funded versus the
grassroots.

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Keynote \note{Lead: S.-C. Choi.  See \href{http://tinyurl.com/qbbqgsj}{slides} \& \href{http://tinyurl.com/q45kfcn}{video}.}} \label{sec:keynote} 
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

Turk reminded the audience that the volunteer model means unpaid labor. On this
note, he recommended Ashe Dryden's blog post on the
\href{http://www.ashedryden.com/blog/the-ethics-of-unpaid-labor-and-the-oss-community}{ethics
of unpaid labor and the open source software community}. Often times, a person
funded to work full time on some scientific project can spend a small amount of
time for working on a piece of software necessary for the project. However, all
researchers' ability to participate in that volunteer community are not always
the same and may not always be aligned with the research projects. From Turk's
experience, we cannot always rely on unpaid labor and volunteer time to sustain
a piece of software---this came down to the notions of the top-down and the
bottom-up approaches, i.e., the funded versus the grassroots. However, Turk
pointed out that bottom-up volunteers driven projects can be just as large scale
as a top-down model software development.

When Turk talked about sustaining scientific software, it really meant to him as
conducting scientific inquiries by often some specific software and sustaining
the people we care about, our careers, and the future of our fields. According
to Turk, we all have an invested stake in sustaining scientific software. Again
it came down to the top-down versus bottom-up notion that we can have anointed
specific software packages to be sustained at the expense of new grassroot
developments.

To close, Turk talked about possibilities of why we want to sustain scientific
software: devotion to science and interests in pursuing the next stage of
research; fun and creative thrill in writing code and paper; usefulness with
measurable impacts, for example, LINPACK and HDF groups providing of data
storage to satellites, which goes beyond usefulness to necessity. Lastly, Turk
presented his wishlist of questions to be answered: 
%
\begin{itemize}

\item How do we ship a product on time when dealing with a mix of funding models
and motivations especially when we rely on volunteers?

\item how do we know when it is time to end some software and move on? For
example, shall we stop sustaining Python and switch to Julia and Javascript?

\item How can productized software balance its future versus its past, or the new
needs of the customers versus the existing needs of the development community?

\item How can we help avoid burnout and retain the joy in the communities?

\item How can we reduce systemic bias, which goes back to the blog post of
Dryden especially on how ethics of unpaid labor disproportionately affect
underrepresented communities?

\end{itemize}
